---
title: "Statistical analyses of post - pre training differences"
author: Tyler Sagendorf
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Statistical analyses of post - pre training differences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
# Required packages
library(MotrpacRatTraining6moWATData)
library(ggplot2)
library(dplyr)
library(emmeans)
library(tibble)
library(tidyr)
library(purrr)
theme_set(theme_bw()) # base plot theme
```

```{r}
# Generate all regression diagnostic plots
plot_lm <- function(x) {
  par(mfrow = c(2, 3))
  for (i in 1:6)
    plot(x, which = i)
  par(mfrow = c(1, 1))
}
```


# Regression Models

We will fit OLS or WLS regression models with sex, timepoint, and their interaction as predictors of the (post - pre) training differences for each phenotypic measure. Only those samples selected for -omics analysis will be considered. If a WLS model is used, the weights will be reciprocal group variances.


## NMR Weight

```{r, fig.height=4, fig.width=5}
# Plot points
ggplot(PHENO_WAT, aes(x = timepoint, y = post_weight - pre_weight,
                      color = omics_analysis)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.4),
             na.rm = TRUE, alpha = 0.5) +
  facet_grid(~ sex, drop = TRUE) +
  labs(y = "NMR weight (post - pre)") +
  scale_color_manual(values = c("red", "black"),
                     breaks = c(TRUE, FALSE)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "bottom", 
        legend.direction = "horizontal")
```

There is one negative outlying value in the SED male group that may affect the model fit. We will start with a WLS model, since we observe different group variances.

```{r}
wt.weight <- filter(PHENO_WAT, omics_analysis) %>% 
  group_by(sex, timepoint) %>% 
  mutate(1 / var(post_weight - pre_weight, na.rm = TRUE)) %>% 
  pull(-1)

fit.weight <- lm(I(post_weight - pre_weight) ~ sex * timepoint,
                 weights = wt.weight,
                 data = filter(PHENO_WAT, omics_analysis))
plot_lm(fit.weight)
```

The diagnostic plots appear relatively fine. We will try to simplify the model.

```{r}
anova(fit.weight, test = "F")
```

All terms are significant. We will not change the model.

```{r}
summary(fit.weight)
```


## NMR Fat Mass

```{r, fig.height=4, fig.width=5}
# Plot points
ggplot(PHENO_WAT, aes(x = timepoint, y = post_fat - pre_fat,
                      color = omics_analysis)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.4),
             na.rm = TRUE, alpha = 0.5) +
  facet_grid(~ sex, drop = TRUE) +
  labs(y = "NMR Fat (post - pre)") +
  scale_color_manual(values = c("red", "black"),
                     breaks = c(TRUE, FALSE)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "bottom", 
        legend.direction = "horizontal")
```

We observe unequal group variances, and an outlying value in the 4W males. We will fit a WLS model with reciprocal group variances as weights.

```{r}
wt.fat <- filter(PHENO_WAT, omics_analysis) %>% 
  group_by(sex, timepoint) %>% 
  mutate(1 / var(post_fat - pre_fat, na.rm = TRUE)) %>% 
  pull(-1)

fit.fat <- lm(I(post_fat - pre_fat) ~ sex * timepoint,
              weights = wt.fat,
              data = filter(PHENO_WAT, omics_analysis))
plot_lm(fit.fat)
```

The diagnostic plots seem mostly fine. We will try to simplify the model.

```{r}
anova(fit.fat, test = "F")
```

All terms are significant.

```{r}
summary(fit.fat)
```


## NMR % Fat

```{r, fig.height=4, fig.width=5}
# Plot points
ggplot(PHENO_WAT, aes(x = timepoint, y = post_fat_pct - pre_fat_pct,
                      color = omics_analysis)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.4),
             na.rm = TRUE, alpha = 0.5) +
  facet_grid(~ sex, drop = TRUE) +
  labs(y = "NMR % Fat (post - pre)") +
  scale_color_manual(values = c("red", "black"),
                     breaks = c(TRUE, FALSE)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "bottom", 
        legend.direction = "horizontal")
```

We observe the same outlying value in the 4W male group. We will fit an OLS model first.

```{r}
fit.fat_pct <- lm(I(post_fat_pct - pre_fat_pct) ~ sex * timepoint,
                  data = filter(PHENO_WAT, omics_analysis))
plot_lm(fit.fat_pct)
```

We will incorporate reciprocal group variances as weights.

```{r}
wt.fat_pct <- filter(PHENO_WAT, omics_analysis) %>% 
  group_by(sex, timepoint) %>% 
  mutate(1 / var(post_fat_pct - pre_fat_pct, na.rm = TRUE)) %>% 
  pull(-1)

fit.fat_pct <- update(fit.fat_pct, weights = wt.fat_pct)
plot_lm(fit.fat_pct)
```

The diagnostic plots look better, though still not great. We will try to simplify the model.

```{r}
anova(fit.fat_pct, test = "F")
```

All terms are significant, so we will not update the model.

```{r}
summary(fit.fat_pct)
```


## VO$_2$max Relative to Body Mass

We will look at VO$_2$max ($\text{mL} \cdot (\text{kg body mass})^{-1} \cdot \text{min}^{-1}$).

```{r, fig.height=4, fig.width=5}
# Plot points
ggplot(PHENO_WAT, aes(x = timepoint, 
                      y = post_vo2max_ml_kg_min - pre_vo2max_ml_kg_min,
                      color = omics_analysis)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.4),
             na.rm = TRUE, alpha = 0.5) +
  facet_grid(~ sex, drop = TRUE) +
  labs(y = "Absolute VO2max (post - pre)") +
  scale_color_manual(values = c("red", "black"),
                     breaks = c(TRUE, FALSE)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "bottom", 
        legend.direction = "horizontal")
```

The SED female group has a large outlying value. We will begin by fitting an OLS model.

```{r}
fit.vo2max_rel <- lm(I(post_vo2max_ml_kg_min - pre_vo2max_ml_kg_min) ~ 
                       sex * timepoint,
                     data = filter(PHENO_WAT, omics_analysis))
plot_lm(fit.vo2max_rel)
```

We will incorporate reciprocal group variances as weights.

```{r}
wt.vo2max_rel <- filter(PHENO_WAT, omics_analysis) %>% 
  group_by(sex, timepoint) %>% 
  mutate(1 / var(post_vo2max_ml_kg_min - pre_vo2max_ml_kg_min, 
                 na.rm = TRUE)) %>% 
  pull(-1)

fit.vo2max_rel <- update(fit.vo2max_rel, weights = wt.vo2max_rel)
plot_lm(fit.vo2max_rel)
```

The diagnostic plots look mostly fine. We will try to simplify the model.

```{r}
anova(fit.vo2max_rel, test = "F")
```

Only `timepoint` appears significant.

```{r}
fit.vo2max_rel.1 <- update(fit.vo2max_rel, formula = . ~ timepoint)
anova(fit.vo2max_rel.1, fit.vo2max_rel, test = "F")
```

There is no significant difference between the models, so we will use the simpler one.

```{r}
fit.vo2max_rel <- fit.vo2max_rel.1
plot_lm(fit.vo2max_rel)
```

The diagnostic plots look mostly fine.

```{r}
summary(fit.vo2max_rel)
```


## VO$_2$max Relative to Lean Mass

We will look at VO$_2$max ($\text{mL} \cdot (\text{kg lean mass})^{-1} \cdot \text{min}^{-1}$).

```{r, fig.height=4, fig.width=5}
# Plot points
ggplot(PHENO_WAT, aes(x = timepoint, 
                      y = post_vo2max_ml_kg_lean_min - pre_vo2max_ml_kg_lean_min,
                      color = omics_analysis)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             jitter.height = 0, 
                                             dodge.width = 0.4),
             na.rm = TRUE, alpha = 0.5) +
  facet_grid(~ sex, drop = TRUE) +
  labs(y = "Relative VO2max (post - pre)") +
  scale_color_manual(values = c("red", "black"),
                     breaks = c(TRUE, FALSE)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "bottom", 
        legend.direction = "horizontal")
```

There is a large outlying value in the SED female group, though it is comparable to observations of samples not selected for -omics analyses. We will fit a WLS model with reciprocal group variances as weights.

```{r}
wt.vo2max_rel_lean <- filter(PHENO_WAT, omics_analysis) %>% 
  group_by(sex, timepoint) %>% 
  mutate(1 / var(post_vo2max_ml_kg_lean_min - pre_vo2max_ml_kg_lean_min, 
                 na.rm = TRUE)) %>% 
  pull(-1)

fit.vo2max_rel_lean <- lm(I(post_vo2max_ml_kg_lean_min - 
                              pre_vo2max_ml_kg_lean_min) ~ 
                            sex * timepoint,
                          weights = wt.vo2max_rel_lean,
                          data = filter(PHENO_WAT, omics_analysis))
plot_lm(fit.vo2max_rel_lean)
```

The diagnostic plots look fine. We will try to simplify the model.

```{r}
anova(fit.vo2max_rel_lean, test = "F")
```

The interaction is not significant.

```{r}
fit.vo2max_rel_lean <- update(fit.vo2max_rel_lean, 
                              formula = . ~ sex + timepoint)
plot_lm(fit.vo2max_rel_lean)
```

The diagnostic plots look mostly fine. 

```{r}
summary(fit.vo2max_rel_lean)
```


# Comparisons

We will test whether the mean of each timepoint is significantly different from 0 by sex, if sex was included as a predictor. P-values will be adjusted across each set of 3 comparisons by sex using the Holm method.

```{r}
## Create emm_list objects
model_list <- list("NMR Weight" = fit.weight,
                   "NMR Fat Mass" = fit.fat,
                   "NMR % Fat" = fit.fat_pct,
                   "VO2max (relative to body mass)" = fit.vo2max_rel,
                   "VO2max (relative to lean mass)" = fit.vo2max_rel_lean)

# Extract model info
model_df <- model_list %>% 
  map_chr(.f = ~ paste(deparse(.x[["call"]]), collapse = "")) %>% 
  enframe(name = "response", 
          value = "model") %>% 
  mutate(model = gsub("(?<=[\\s])\\s*|^\\s+|\\s+$", "", model, perl = TRUE),
         model_type = sub("^([^\\(]+).*", "\\1", model),
         formula = sub(".*formula = ([^,]+),.*", "\\1", model),
         family = sub(".*family = ([^\\)]+\\)),.*", "\\1", model),
         family = ifelse(model_type == "lm", "gaussian", family),
         # if weights were used, they were reciprocal group variances
         weights = ifelse(grepl("weights = ", model), 
                          "reciprocal group variances", NA)) %>% 
  dplyr::select(-model)

# Results
PHENO_WAT_STATS <- map(model_list, function(mod_i) {
  terms_i <- attr(terms(mod_i), which = "term.labels")
  by <- intersect("sex", terms_i)
  
  if (length(by) == 0) {
    by <- NULL
  }
  
  emmeans(mod_i, specs = "timepoint", by = by, 
          infer = TRUE, adjust = "holm") %>% 
    summary() %>% 
    as.data.frame()
}) %>%
  enframe(name = "response") %>% 
  unnest(value) %>%
  mutate(signif = cut(p.value, 
                      breaks = c(0, 0.001, 0.01, 0.05, 1),
                      labels = c("***", "**", "*", ""),
                      include.lowest = TRUE, right = FALSE,
                      ordered_result = TRUE)) %>% 
  relocate(p.value, .before = signif) %>% 
  left_join(model_df, by = "response") %>% 
  pivot_longer(cols = contains(".ratio"), 
               names_to = "statistic_type", 
               values_to = "statistic", 
               values_drop_na = TRUE) %>% 
  relocate(starts_with("statistic"), .before = p.value)
```

See `?PHENO_WAT_STATS` for details.

```{r}
print.data.frame(head(PHENO_WAT_STATS))
```

```{r, eval=FALSE, include=FALSE}
# Save data
usethis::use_data(PHENO_WAT_STATS, internal = FALSE, overwrite = TRUE,
                  compress = "bzip2", version = 3)
```

# Session Info

```{r}
sessionInfo()
```

