---
title: "LIMMA Diagnostic Plots"
author: "Tyler Sagendorf"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_document:
    toc: true
---

This article generates multidimensional scaling (MDS) and mean--variance trend plots (Extended Data Fig. 8A--G).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.height = 2,
  fig.width = 2.3,
  dpi = 400
)
```

```{r setup}
# Required packages
library(MotrpacRatTraining6moWATData)
library(MotrpacRatTraining6moWAT) # theme_pub, ggplotMDS
library(Biobase)
library(edgeR)
library(dplyr)
library(ggplot2)

save_plots <- dir.exists(paths = file.path("..", "..", "plots"))
```

```{r}
# Transcriptomics pre-processing
# Convert filtered counts to normalized log2 counts-per-million reads
dge <- DGEList(counts = exprs(TRNSCRPT_EXP),
               samples = pData(TRNSCRPT_EXP),
               group = TRNSCRPT_EXP$exp_group)
dge <- calcNormFactors(dge, method = "TMM")
m_dge <- TRNSCRPT_EXP
exprs(m_dge) <- cpm(dge, log = TRUE)
```


# Multi-dimensional Scaling Plots

## PROT

```{r}
mds_prot <- ggplotMDS(PROT_EXP)

mds_prot
```

```{r, eval=save_plots}
ggsave(file.path("..", "..", "plots", "MDS_PROT.pdf"), mds_prot,
       height = 2, width = 2.3, units = "in", dpi = 400, bg = "white")
```

## TRNSCRPT

```{r}
mds_trnscrpt <- ggplotMDS(m_dge)

mds_trnscrpt
```

```{r, eval=save_plots}
ggsave(file.path("..", "..", "plots", "MDS_TRNSCRPT.pdf"), mds_trnscrpt,
       height = 2, width = 2.3,
       units = "in", dpi = 300, bg = "white")
```

## PHOSPHO

```{r}
mds_phospho <- ggplotMDS(PHOSPHO_EXP)

mds_phospho
```

```{r, eval=save_plots}
ggsave(file.path("..", "..", "plots", "MDS_PHOSPHO.pdf"), mds_phospho,
       height = 2, width = 2.3,
       units = "in", dpi = 300, bg = "white")
```

## METAB

```{r}
mds_metab <- ggplotMDS(METAB_EXP)

mds_metab
```

```{r, eval=save_plots}
ggsave(file.path("..", "..", "plots", "MDS_METAB.pdf"), mds_metab,
       height = 2, width = 2.3,
       units = "in", dpi = 300, bg = "white")
```


# Mean--variance trend

```{r}
# Do not run the following chunks when knitting
knitr::opts_chunk$set(eval = FALSE)
```

## PROT and PHOSPHO

```{r}
# PROT and PHOSPHO:
# In vignettes/WATSC_DA.Rmd,
# 1) debug limma_full.  Set 
#    plot = TRUE. Run all code up through the creation of fit.smooth.
# 2) debug limma::plotSA. Run plotSA(fit.smooth) up until the plot is made.
# 3) Set working directory to the location of this article and run the following
#    code. Modify the -ome in the file name as needed (either PROT or PHOSPHO).

ome <- "PROT" # PROT or PHOSPHO

# Create data.frame so that red points (outliers) will be plotted on top
mv_prot_df <- data.frame(x = x,
                         y = y,
                         color = colv,
                         s = sqrt(sqrt(fit$s2.prior))) %>%
  mutate(color = factor(color, levels = c("black", "red"),
                        labels = c("Normal", "Outlier"))) %>%
  arrange(color)

# Plot
mv_prot <- ggplot(mv_prot_df) +
  geom_point(aes(x = x, y = y, color = color),
             size = 1, alpha = 0.5, shape = 16) +
  geom_line(aes(x = x, y = s), linewidth = 0.3,
            color = "red", lty = "dashed") +
  coord_cartesian(ylim = c(0, 3)) +
  scale_y_continuous(expand = expansion(mult = c(5e-3)),
                     limits = c(0, NA)) +
  scale_color_manual(name = NULL,
                     values = c("black", "red"),
                     labels = c("Normal", "Outlier")) +
  labs(x = "Average log-expression",
       y = expression(sqrt("sigma")),
       subtitle = "eBayes(trend=TRUE, robust=TRUE)") +
  theme_pub() +
  theme(axis.line.y.right = element_blank(),
        strip.text = element_text(hjust = 0,
                                  margin = margin(t=0, r=0, b=5, l=0)),
        legend.key.size = unit(6, "pt"),
        # legend.margin = margin(r = 0, l = 0),
        legend.position = c(0.8, 5/6),
        legend.background = element_rect(fill = "white", color = "black",
                                         linewidth = 0.3),
        panel.grid.minor = element_line(linewidth = 0))

ggsave(
  filename = sprintf(file.path("..", "..", "plots", "limma_diagnostics_mean-variance_%s.pdf"), ome),
  mv_prot, height = 2, width = 2.3, dpi = 400, bg = "white"
)
# exit debugger now
```

## TRNSCRPT

```{r}
# In vignettes/WATSC_DA.R:
# 1) debug `limma_full`. Run all code before `limma::voomWithQualityWeights` 
#    is called.
# 2) debug `limma::voomWithQualityWeights`. Set plot=TRUE. Run all code before
#    the second use of `voom`.
# 3) debug `voom`. Run all code before `plot(sx, sy, ...)`
# 4) Run the following code:

mv_trnscrpt <- ggplot(mapping = aes(x = sx, y = sy)) +
  geom_point(alpha = 0.3, size = 1, shape = 16) +
  geom_smooth(formula = y ~ x, method = "loess",
              linewidth = 0.3,
              color = "red", lty = "dashed") +
  labs(x = expression(paste("log"[2],"(count size + 0.5)")),
       y = expression(sqrt("standard deviation")),
       subtitle = "voomWithQualityWeights(..., plot=TRUE)") +
  scale_y_continuous(limits = c(0, 2.5),
                     expand = expansion(mult = 5e-3)) +
  theme_pub() +
  theme(axis.line.y.right = element_blank(),
        strip.text = element_text(hjust = 0,
                                  margin = margin(t=0, r=0, b=5, l=0)),
        legend.key.size = unit(6, "pt"),
        legend.margin = margin(r = 0, l = 0),
        panel.grid.minor = element_line(linewidth = 0))

ggsave(file.path("..", "..", "plots", "limma_diagnostics_mean-variance_TRNSCRPT.pdf"),
       mv_trnscrpt, height = 2, width = 2.3, dpi = 400, bg = "white")
# exit debugger now
```

## METAB

Separate mean--variance trends were fit to each platform, so there would be about a dozen plots. We will leave this as an exercise for the interested reader.

